The Name of the Title is Hope

### 2023-04-28 input/2_introduction.tex

\section{Introduction}
% 1. Describe the Current Situtation
% Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.
% 2. What is the complication, challenge identified
% Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
Technological advances in hardware that allow physically moving a user's body have reignited the dreams of overcoming human limitations, recorvering lost abilities as well as simplifying skill acquisition. Wearable exoskeletons move a user's body by applying forces to the extremities, for example by pulling on the fingertips (see Dexmo Haptic Glove \footnote{https://www.dextarobotics.com/}) while electrical muscle stimulation (EMS) makes the user's extremities move by sending current into the muscle activating nerves. These technologies are becoming increasingly seamless to use. However, augmented user's frequently report dissociative experiences following augmented action. They do not experience agency.

The sense of agency refers to the experience of being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us. This being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ubd} can be subdivided into the \textit{feeling} of agency, a low level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}. The key challenge in human \textit{physical augmentation} is to create an experience of pre-reflective agency, so user's feel as though they are in the "driving seat" once again.
% if something happens for you in an interaction there is a cost to that:
% - impacts sense of agency which in turn makes user's engage less
% - impacts precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control

% What have others done to address this and what do we propose "instead"
% summarize Kasahara's work
This challenge has driven researcher

% 3. Question
% Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
But how can we build a physical augmentation system that keeps user's engaged in the action, preserving sense of agency? In other words how can we design for users to experience natural augmentation?

% 4. (short) Answer Teaser
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
% keep this short
In this paper, we present an augmentation that maintains the user's sense of agency. Our system establishes a fast communication channel between the user's thoughts and a physical end effector, here EMS. The system controls the user's muscles through EMS at the time of the user's intent to interact, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG). In our user study we then investigated whether keeping the physical impact on the user's body in line with their intention to move, preserves their sense of agency.

\subsection{Preserving Agency using Brain Signal of the Intent to Interact}
% 4. (long) Anwser with teaser image
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
% Now extend into a longer paragraph with a teaser image explaining the system in the following subsection

% not only physical integration but also other adpative interfaces, or control of mobile phone

\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm.}
\input{input/1_teaser_figure}

>>>>

\section{Introduction}
% 1. Describe the Current Situtation
% Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.
% 2. What is the complication, challenge identified
% Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.

Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities as well as simplifying skill acquisition~\citep{Kasahara2019-sk, Goto2020-mw}. There are two main technologies to physically augment users' actions: First, through the use of mechanical actuators, exoskeletons, a user's body can be moved by applying forces to the extremities, for example by pulling on the fingertips, and second, by making the user's extremities move by sending current into their muscle activating nerves using electrical muscle stimulation (EMS). However, augmented user's frequently report dissociative experiences during or following augmented action~\citep{}. They do not experience agency.
\todo{this needs a good citation}

The experience of agency means being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us. It has frequently been shown that if users have a strong sense of agency, they are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system~\citep{Berberian2012-do, Miller2007-rb}. The key challenge to drive adoption of human \textit{action augmentation} then is to design for agency experience, so user's feel as though they are in the ``driving seat'' once again.
% if something happens for you in an interaction there is a cost to that:
% - impacts sense of agency which in turn may makes user's engage less -> citation?
% - may impact precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control -> citation?

% 3. Question
% Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
% What have others done to address this and what do we propose "instead"
Previously, ~\citet{Kasahara2019-sk} have shown that reaction times when tapping targets on a touchscreen can be increased by augmenting the user through the use of EMS while maintaining their sense of agency. The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e. how much earlier the action of the user can be elicited. However, setting this gain factor may depend on each specific task and requires fitting the parameters over and over again. A key question then remains: How can we maintain agency during physical action augmentation without any training or running the risk of over-fitting task specific parameters? In other words, how can we design closed systems without the need for external tuning, to deliver an integrated, \textit{natural}, experience of user's augmented actions?
% At around 80 ms preceding the naturally occurring action, user's integrate the augmentation and maintain an experience of agency.

\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm. <- better use this for the teaser image. Here better present the main results, i.e. sense of agency scores?}
\input{input/1_teaser_figure}

% 4. (short) Answer Teaser
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
% keep this short
In this paper, we present a fast communication channel between the user's brain signals and a physical end effector, here EMS. Crucially, our system controls the user's muscles through EMS at the time of the user's \textit{intent to act}, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG). In our user study we applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their sense of agency.
% We found ... rendering the augmentation a \textit{natural}, integrated expansion of the user's body.

% \subsection{Preserving Agency using Brain Signals of the Intent to (Inter)act}
% 4. (long) Anwser with teaser image
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.

### 2023-04-28 input/3_related_work.tex

We present a bodily integration experience moving the finger of a user after the user intends to do so, but before the user directs its muscles to move.

\subsection{The Disruption in Sense of Agency in Augmented Humans}

\subsection{Assessing the Sense of Agency when Integrating with Computers}
% how can we measure our outcome measure sense of agency?
% - questionnaires
% - intentional binding
% - phenomenological interview

\subsection{Brain-Computer Interfaces for Intent Prediction}
% Single-trial Classification of the Intent to Interact
Our idea to overcome the disruption in agency is to use EEG signals reflecting (inter)action intent. One such signal is the Readiness Potential (RP), or \textit{lateralized readiness potential}. It is an amplitude fluctuation that has been frequently observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}. It is reliably observed in electrodes placed over the motor cortex contralateral to the acting hand. Since its measurable onset precedes the time of participants self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective. However, evidence abounds for its role in action preparation. An RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement. The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after. The second stage manifests contra-laterally in the primary motor cortex~\cite{Shibasaki2006-mt}. A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}. In their study,~\cite{Schultze-Kraft2021-cu} asked participants to find a way to perform a voluntary movement so that the accompanying RP amplitude was as small as possible. After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for. They found participants \textit{not} to be able to suppress their RP. This inability to suppress the RP renders it a reliable feature for classification. While its role in movement preparation remains somewhat controversial, it is established as a signal occurring during movement planning and motivates a usage as a reliable predictor of movement.

>>>>

% In this paper, we present a physical action augmentation prototype and a user study to investigate the experience of agency during the use of the prototype. Hence, our research builds on previous qualitative and quantitative research on the sense of agency.
% on actuated haptic systems, brain-computer interfacing as well as
% moving the finger of a user when the user intends to do so, but before the user directs their muscles to move.

The sense of agency, or this being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ub} can be subdivided into the \textit{feeling} of agency, a low level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}.

\subsection{Measuring agency experience}
% explicit measures - reflective (short paragraph)
% - questionnaires
% - phenomenological interview

% implicit measures - pre-reflective (longer paragraph)
% - intentional binding: explain what this is and give a bit more detail about the underlying mechanisms and ideas about whats going on in the brain here

\textit{Intentional binding} is the phenomenon of subjectively compressing the time interval between a voluntary action and the sensory consequences of that action~\cite{Moore2012-ic}.

% what does that have to do with agency?
\subsection{Sense of agency and actuated haptic systems}

% close this paragraph with the problem that of the control mechanism of the

### 2023-04-28 input/4_prototype.tex

\section{Methods}

With this study we wanted to find out whether agency can be preserved during \textit{physical actuation} of the user's body when their own brain signals serve as the control signal. To this end, we build a system that leverages the brain signals known to reflect the volitional intent to interact, i.e. the EEG readiness potential, to trigger electrical muscle stimulation to cue a reaching movement.

To assess if agency was preserved, we used a mixed methods approach including a psychometric test of intentional binding, questionnaires as well as qualitative interviews. Our interviews followed a phenomenological structure \todo{add reference from AH review paper mueller}. We t

\subsection{Participants}

\todo{add actual number} Ten participants (\todo{add mean age and sd}) from our local institution participated in our study. All participants were right handed, had normal or corrected to normal vision and had not experienced EMS before. Participants were compensated with 12 Euro per hour of study participation. Prior to their participation, participants were informed of the nature of the experiment, recording and anonymization procedures and signed a consent form.

\subsection{Apparatus}
The experimental setup, depicted in Figure~\ref{}, comprised: (1) a wrist-mounted wearable HTC VIVE tracker for motion capture, (2) a 64-channel EEG system, and (3) a medically-compliant EMS device connected via two electrodes worn on the forearm. To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code to the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.

\indent\textbf{(1) Hand Tracking.} We used an HTC Vive Tracker, attached to the participant's wrist, to track their right hand at 90Hz. Therefore, we ran a simple Unity3D scene and streamed the data via the labstreaminglatyer (LSL)\footnote{https://github.com/sccn/labstreaminglayer}.

\indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts. Electrodes were placed according to the extended international 10–20 system \cite{Chatrian1985-ys}. One electrode was placed under the left eye to provide additional information about eye movements (vEOG). After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity and electrode impedance was brought below 10k$\Omega$ for all electrodes. EEG data was recorded with a sampling rate of 250 Hz. We used LSL to make the EEG data stream available in the network and synchronize the recordings of EEG data, motion capture and an experiment marker stream that marked sections of the study procedure.

\missingfigure{show participant with measurement system and description next to each element}

\indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the index finger via electrical muscle stimulation (EMS), which was delivered via two electrodes attached to the participants' extensor digitorum muscle. We utilized the extensor digitorum since we found that we can robustly actuate it without inducing parasitical motion of neighboring muscles. This finger actuation was achieved via a medically-compliant battery powered muscle stimulator (prorelax TENS/EMS Super Duo Plus). The EMS system's output was controlled by flipping a solid state relay connected via an Arduino Uno to the experiment computer. The EMS was pre-calibrated per participant to ensure a pain-free stimulation and robust actuation.

\subsection{Task}
\missingfigure{add an infographic of the task progression with a time axis here, this should be covering two columns and be at the top of the page!}
Participants performed a simple reaching task: resting their hand on the table, participants were instructed to press a specified key on a notebook in front of them. For each trial, a key was randomly selected from one of two groups of letters: (1) 'a', 's', 'd' or (2) 'j', 'k', 'l'. The selected group was then changed with every trial. This was done to keep the task engaging while maintaining a comparable reaching movement between trials.

The interaction flow for one trial of our task, depicted in Figure~\ref{task_flow}, was as follows: (1) participants waited for a new letter to appear on screen; (2) then, they were instructed to wait for a brief moment (~2s), before (3) reaching out and pressing the key. In line with the literature on the origin of the RP generating process they were told "to avoid preplanning the movement, avoid any obvious rhythm, and to press when they felt the spontaneous urge to move". (4) Two \todo{how many, but do keep this constant in order for them to focus on the duration of their action} seconds after they pressed the key, they were presented with a tone and asked to judge whether their movement, from the start of their reach to the button press, lasted longer or shorter than the tone. (5) When they perceived the tone to be longer, they had to press the up key or vice versa the down key for perceiving that the tone was shorter than their movement. Following an inter-trial-interval the next trial started.

\subsubsection{Experiment Design and Procedure}
Participants completed three experimental blocks. The first block consisted of 60 trials and was used to obtain training data for the single-trial classification system outlined below.

%The condition of the second and third block, see below, were alternated between participants.

In a second block, participants completed 80 trials of the same task and trial order as in the first block. Now, the brain-computer interface (BCI) detecting readiness potentials from real-time EEG data was continuously running in the background. After the target letter was shown to the participants and they were preparing to reach out and press the corresponding key, the BCI was live. Whenever the BCI detected EEG signals to reflect an interaction intent, EMS was activated causing the participants index finger to lift.

Both experimental blocks two and three were followed by a mixed methods data collection of participants' perceived agency.

In summary, the experiment consisted of four phases: (1) a setup phase; (2) a first experimental block: the training phase; (3) a second experiment block: the test phase; and (4) a test battery of questionnaire and interview to inquire about their experienced agency.

@@

\subsection{Interview \& Questionnaire}

To assess if agency was preserved, three measures were recorded during and after the test phase.

\indent\textbf{(1) Intentional Binding.}
At the end of each trial participants were tasked to compare the duration of their arm movement to a tone and indicate whether the tone was longer or shorter. We set the initial duration at 500 ms for each participant. When participants indicated the tone to be longer or shorter by pressing either the up or down key we added, or respectively subtracted, 100 ms from the tone duration. This new duration was then used for the next trial and so on. With 60 trials in the training phase, we hypothesized participants would oscillate 50-50 around a target duration after about 2/3 of the training trials. The final duration after all training trials was then used again as the start duration during the test phase. In line with the literature on intentional binding, we hypothesized this duration to remain stable in the case that agency was preserved during the test phase with EMS. A deviation of the tone duration from the training phase was hypothesized to reflect a disruption in experienced agency.

\indent\textbf{(2) Questionnaire.}
In line with~\cite{Hornbaek} we prompted participants with three statements and a 7-point Likert scale centered at 0. The first one ``I felt like the device I was using was part of my body'' was in relation to body ownership. The following two statements ``It felt like I was in control of the movements during the task.'' and ``What is the degree of control you felt?'' prompted participants on their experienced agency.

\indent\textbf{(3) Interview.}
We interviewed user's about the experience with the EMS integration. We asked open how questions in a collaborative interview inquiring about the user's lived experience. After prompting user's to recall their experience we started the interview with: "how did it feel like?". If user's made reference themselves to the EMS device we followed up by inquiring: "why do you think it (the EMS trigger) happened?", "What do you think triggered it?". The goal was to learn about the source attribution user's made about the EMS trigger. Since we leveraged the RP, we hypothesized that user's would attribute, at least in part, the EMS trigger to themselves, i.e. their movement intention. Next, we pressed deeper into learning about the cooperative nature between user and system. We asked "Do you believe you did it alone?" and when they made reference to a specific situation where the EMS was triggered we followed up with "What did it feel like, did you had the feeling the system controlled you or did you cooperate?" and "What would have to be there for it to feel like a cooperation?". We closed the interview with two additional exploratory questions: "Did the system influence your performance? If yes, how and why?" as well as asking them about any application scenarios they have in mind.

To not break the user's immersion by disturbing the flow of the interaction we decided against asking qualitative questions at the end of each and every trial.

\subsection{Statistical Analyses}
\todo{how where the results analyzed: sentiment analyses and statistics of IB data and questionnaire results}

% ERP with significance mask for Cz? Assessing significance of classifier

% IB Task: ttest of final ib time between with and without EMS -> # subjects input values
% did they increase the IB task in more than 50% of the trials after EMS?

% questionnaires: ttest against zero to report direction

% interviews: anecdotal summary and sentiment analyses with two groups splitting questionnaires at 0

% correlation with EEG classifier accuracy?

% evaluation of EEG classifier: can report false negatives, how often was it missing to predict a movement onset? hand was moving without EMS being triggered occurred how often? Not possible to evlaute false positives as no ground truth available

\subsection{Reproducing Results and Data Availability}
Data, experimental protocol, analyses code including scripts for a reproduction of the presented results are hosted at open science foundation (OSF)\footnote{[anonymized]}. BIDS formatted data is hosted on openneuro \cite{}.

>>>>

\subsection{Controlling actuated haptic experiences using brain signals reflecting the intent to interact}

To discriminate between the two \textit{action} states, we leveraged the Readiness Potential (RP), or \textit{lateralized readiness potential}. It is an amplitude fluctuation that has been frequently observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}. It is reliably observed in electrodes placed over the motor cortex contralateral to the acting hand. Since its measurable onset precedes the time of participants self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective. However, evidence abounds for its role in action preparation. An RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement. The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after. The second stage manifests contra-laterally in the primary motor cortex~\cite{Shibasaki2006-mt}. A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}. In their study,~\cite{Schultze-Kraft2021-cu} asked participants to find a way to perform a voluntary movement so that the accompanying RP amplitude was as small as possible. After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for. They found participants unable to suppress their RP. This inability to suppress the RP renders it a reliable feature for classification.

% While its role in movement preparation remains somewhat controversial, it is established as a signal occurring during movement planning and motivates a usage as a reliable predictor of movement.

% \subsection{Action augmentation}
% TODO shorten this down to only the 'prototype' part -> EEG controlling the arduino switch. make obvious that what is attached to the switch can be different -> here in this particular study we used EMS -> see User Study

% \subsubsection{Apparatus}
% The experimental setup, depicted in Figure~\ref{}, comprised: (2) a 64-channel EEG system, and (3) a medically-compliant EMS device connected via two electrodes worn on the forearm. To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code to the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.

% \indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts. Electrodes were placed according to the extended international 10–20 system \cite{Chatrian1985-ys}. One electrode was placed under the left eye to provide additional information about eye movements (vEOG). After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity and electrode impedance was brought below 10k$\Omega$ for all electrodes. EEG data was recorded with a sampling rate of 250 Hz. We used LSL to make the EEG data stream available in the network and synchronize the recordings of EEG data, motion capture and an experiment marker stream that marked sections of the study procedure.

% \missingfigure{show participant with measurement system and description next to each element}

% \indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the index finger via electrical muscle stimulation (EMS), which was delivered via two electrodes attached to the participants' extensor digitorum muscle. We utilized the extensor digitorum since we found that we can robustly actuate it without inducing parasitical motion of neighboring muscles. This finger actuation was achieved via a medically-compliant battery powered muscle stimulator (prorelax TENS/EMS Super Duo Plus). The EMS system's output was controlled by flipping a solid state relay connected via an Arduino Uno to the experiment computer. The EMS was pre-calibrated per participant to ensure a pain-free stimulation and robust actuation.

@@
